Fission product decay chain equilibria in \zirconia

### 2019-02-12 manuscript.tex
Update abstract

As fission products proceed down the decay chain from Te to Cs, the defects they produce in \zirconia\ have progressively larger volumes. At the oxygen site following fission product implantation, Te in a 0 oxidation state will form the defect \ch{Te_{O}^{**}} as the dominant anion site defect, with a defect volume relative to the non-defective crystal of +2.74 \r{A}$^{3}$. These Te nuclei  One of the largest changes in defect volume occurs when \ch{Te_{Zr}^{'''}} decays and produces \ch{I_{Zr}^{'''}} (+27.14 \r{A}$^{3}$). This may then decay and produce the defect \ch{Xe_{Zr}^{''''}}, which leads to an even larger change in defect volume (+28.43 \r{A}$^{3}$).

@@

The majority of thermal fission events occur in the outer rim of a fuel pellet. Given a fission product penetration depth of up to 8 $\mu$m in \zirconia\ \cite{degueldre2001behaviour}, a large proportion of fission products implant within the oxide. It has also been observed in PCI failures, that there is a time delay on the order of 10 minutes between the power ramping of the reactor and the subsequent breach of the cladding \cite{bergenlid1980experimental, wood1983effects, pankaskie1981mechanistic}. This time delay is in line with the half lives of many Te and I isotopes (Table \ref{table:decaydata}) such that significant proportions of these isotopes will decay during this time delay, hinting that these phenomena could be related. \\
%In addition, it is known that irradiated cladding containing fresh fuel does not crack when power ramped  \cite{wood1983effects}. \\

@@

The defect volumes of \ch{Cs_{O}^{**}} and \ch{Cs_{Zr}^{'''}} were 4.38 and 90.37 \r{A}$^{3}$ respectively, less than the Xe defects on the same site. This is consistent with the fact that Cs has a much smaller first ionisation energy than Xe, therefore presenting a smaller energy barrier to reducing it's size on either site to improve lattice fit. Because of this, we do not expect further stresses being imposed on the lattice after Xe decays to Cs. We can therefore consider only Te, I and Xe to be significant for this mechanism, as the only contribution to lattice stresses due to Cs would have occurred by direct implantation after fission.

@@

As fission products proceed down the decay chain from Te to Xe, the defects they produce in \zirconia\ have progressively larger volumes. One of the largest changes in defect volume occurs when \ch{Te_{Zr}^{'''}} decays and produces \ch{I_{Zr}^{'''}} (+27.14 \r{A}$^{3}$). This may then decay and produce the defect \ch{Xe_{Zr}^{''''}}, which leads to an even larger change in defect volume (+28.43 \r{A}$^{3}$). This process may be a contributing factor to fuel cladding failures observed after a power ramp, with reported time delays between ramp and failure aligning with the decay rate of Te and I isotopes.

>>>>

PCI failures in nuclear fuel claddings occur after power ramps, typically with a time delay of tens of minutes. The exact mechanism of failure is not known, however, it is widely believed that fission product SCC is involved in the process to some extent. The internal oxide layer of zirconium-based fuel claddings is protective, and so this layer must be breached before corrosive species can attack the metal underneath. Iodine is one of the main corrosive fission products produced in the fuel, though it is also produced in significant quantities by the decay of Te. Many Te fission product isotopes have half-lives under an hour and high fission yields, making them a large contributor to the iodine inventory during power ramps. Iodine nuclei are also unstable and may decay during the time delay into Xe, with the decay chain continuing to Cs over longer time spans. Fission predominantly occurs at the rim of the fuel pellet near the cladding, leading to fission product implantation within the oxide layer. It is important to determine the effect this can have on the oxide's protective capacity. We therefore examine the defect equilibria and report defect volumes of the fission products Te, I, Xe and Cs in the internal oxide layer of fuel cladding using a quantum mechanical modelling approach.

%As fission products proceed down the decay chain from Te to Xe, the defects they produce in \zirconia\ have progressively larger volumes. The change in defect volume when \ch{Te_{Zr}^{'''}} decays and produces \ch{I_{Zr}^{'''}} is +27.14 \r{A}$^{3}$. This then decays and produces the defect \ch{Xe_{Zr}^{''''}}, which leads to an even larger change in defect volume of +28.43 \r{A}$^{3}$. Similarly, at the oxygen site, \ch{Te_{O}^{**}} decays to produce \ch{I_{O}^{*}}, with a resulting defect volume change of +27.16 \r{A}$^{3}$. The lattice mismatch of these defects in \zirconia\ will generate stresses and promote crack formation, exposing Zr metal. If enough fission products are implanted simultaneously, the rate of crack formation may become faster than new passivating oxide can be produced, and typical I-SCC mechanisms will take over to failure. This process may be a contributing factor to fuel cladding failures observed after a power ramp, with reported time delays between ramp and failure aligning with the decay rate of Te and I isotopes.

@@

The majority of thermal fission events occur in the outer rim of a fuel pellet. Given a fission product penetration depth of up to 8 $\mu$m in \zirconia\ \cite{degueldre2001behaviour}, a large proportion of fission products implant within the oxide. It has also been observed in PCI failures that there is a time delay on the order of 10 minutes between the power ramping of the reactor and the subsequent breach of the cladding \cite{bergenlid1980experimental, wood1983effects, pankaskie1981mechanistic}. This time delay is in line with the half-lives of many Te and I isotopes (Table \ref{table:decaydata}) such that significant proportions of these isotopes will decay during this time delay, hinting that these phenomena could be related. \\
%In addition, it is known that irradiated cladding containing fresh fuel does not crack when power ramped  \cite{wood1983effects}. \\

@@

The defect volumes of \ch{Cs_{O}^{**}} and \ch{Cs_{Zr}^{'''}} were 4.38 and 90.37 \r{A}$^{3}$ respectively, less than the Xe defects on the same site. This is consistent with the fact that Cs has a much smaller first ionisation energy than Xe, therefore presenting a smaller energy barrier to reducing it's size on either site to improve lattice fit. Because of this, we do not expect further stresses being imposed on the lattice after Xe decays to Cs. We therefore consider that only Te, I and Xe will be significant in this mechanism, as the only contribution to lattice stresses due to Cs would have occurred through direct implantation after fission.

@@

As fission products proceed down the decay chain from Te to Xe, the defects they produce in \zirconia\ have progressively larger volumes. The change in defect volume when \ch{Te_{Zr}^{'''}} decays and produces \ch{I_{Zr}^{'''}} is +27.14 \r{A}$^{3}$. This then decays and produces the defect \ch{Xe_{Zr}^{''''}}, which leads to an even larger change in defect volume of +28.43 \r{A}$^{3}$. Similarly, at the oxygen site, \ch{Te_{O}^{**}} decays to produce \ch{I_{O}^{*}}, with a resulting defect volume change of +27.16 \r{A}$^{3}$. The lattice mismatch of these defects in \zirconia\ will generate stresses and promote crack formation, exposing Zr metal. If enough fission products are implanted simultaneously, the rate of crack formation may become faster than new passivating oxide can be produced, and typical I-SCC mechanisms will take over to failure. This process may be a contributing factor to fuel cladding failures observed after a power ramp, with reported time delays between ramp and failure aligning with the decay rate of Te and I isotopes.
